---
title: Differential assessment of systemic and local cytokine levels in patients with
  atherosclerosis
author: "George Rusakovich"
date: "2023-05-11"
output: html_document
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file("D:/Workflow/Atherosclerosis lab/Bloody plaques/"))
set.seed(144)
```

## Libraries

```{r packages, include = FALSE}
library(tidyverse)
library(openxlsx)
library(readxl)
library(ggplot2)
library(psych)
library(ggpubr)
library(corrplot)
library(ctrlGene)
library(NormqPCR)
library(generefer)
library(Hmisc)
library(rstatix)
library(reshape)
library(ggplot2)
```

## Sources

```{r sources}
data <- read_xlsx("Imputated_data.xlsx")
data_st <- read_xlsx("Impated for sign test.xlsx")
raw <- read_xlsx("raw.xlsx")

#
colnames(raw)[colnames(raw) == "TNFb"] <- "TNF-b"
colnames(data)[colnames(data) == "TNFb"] <- "TNF-b"
colnames(data_st)[colnames(data_st) == "TNFb"] <- "TNF-b"
```

## Analysis
### 1. Excluding subjects with unfavorable cardiovascular events (TIA, ONMC);
### 2. Choosing cytokines with detectible concentration in all samples;
### 3. Implementing geNorm algorithm;
### 4. Ploting average M-value for reference cytokines;
### 5. Calculating the pairwise variation between the control cytokines; 

```{r, include=T}

# 1. Exclude subjects with unfavorable cardiovascular events (TIA, ONMC)
raw <- raw %>%
  dplyr::filter(`№ пациента` != 129
                & `№ пациента` != 135
                & `№ пациента` != 145
                & `№ пациента` != 146
                & `№ пациента` != 148)

raw[,c(5:45)] <- sapply(raw[,c(5:45)], as.numeric)

colnames(raw)[1:4] <- c("Sample", "Group", "Plate", "Standard")

raw <- raw %>%
  mutate(Group = ifelse(Group == 1, "Plasma", "Media"))

# 2. Choosing cytokines with detectible concentration in all samples;
na_cols <- which(colSums(is.na(raw)) == 0)
print(na_cols)
test <- raw[,na_cols]
df <- test[,c(1,5:9)]

# 3.  Implementing geNorm algorithm;
results <- ctrlGene::geNorm(df[,c(2:6)],  ctVal=F)

# 4. Ploting average M-value for reference cytokines;
results

#tiff("plotM_NF.tiff", units="in", width=6, height=5, res=800, pointsize = 7, family = "serif")
#jpeg("plotM_NF.jpeg", units="in", width = 6, height = 5, res = 800, family = "serif", pointsize = 7)
plotM(results)
#dev.off()

# 5. Calculating the pairwise variation between the control cytokines;
# Create a matrix of cytokines concentration values
df_raw <- data.frame(
  Sample = c(raw$Sample),
  Control1 = c(raw$`PDGF-AA`),
  Control2 = c(raw$`MIP-1b`),
  Control3 = c(raw$`TNF-a`),
  Control4 = c(raw$`IP-10`),
  Control5 = c(raw$MDC))

df_raw_mat <- as.matrix(df_raw[,2:6])

# 
pairwiseV(df_raw_mat, ctVal = T) 

#tiff("plotV_NF.tiff", units="in", width=5, height=5, res=800, pointsize = 8, family = "serif")
#jpeg("plotV_NF.jpeg", units="in", width = 5, height = 5, res = 800, family = "serif", pointsize = 8)
plotV(pairwiseV(df_raw_mat, ctVal = F))
#dev.off()

# To determine the optimal number of genes for normalization, you want to choose the smallest number of control genes that results in a low pairwise variation. A good rule of thumb is to choose the smallest number of genes that still achieves a low pairwise variation. A commonly used threshold is 0.15 or 0.2, but the optimal threshold can depend on the specific dataset and analysis.

# Vandesompele J, De Preter K, Pattyn F, et al Accurate normalization of real-time quantitative RT-PCR data by geometric averaging of multiple internal control genes. Genome Biol, 3: RESEARCH0034 2002.

# *Note
#  Taking all this into consideration, we recommend the minimal use of the three most stable internal control genes for calculation of an RT-PCR normalization factor (NF n , n = 3), and stepwise inclusion of more control genes until the (n + 1)th gene has no significant contribution to the newly calculated normalization factor (NFn + 1). 

# 
```

### 6. Calculating normalization factor 
### 7. Ploting correlation plot for different normalization factors calculated by different count of reference cetokines
### 8. Comparison of reference cytokines concentration after normalization by different number of reference cytokines included into NF. 

```{r, include=T}

### 6. Calculating normalization factor

colnames(raw)

nf <- raw[,na_cols]

nf <- nf %>%
  mutate(NF_TNFa = `TNF-a`,
         NF_IP10 = `IP-10`,
         NF2 = ((`TNF-a`*`IP-10`)^(1/2)),
         NF3 = ((`TNF-a`*`IP-10`*`MDC`)^(1/3)),
         NF4 = ((`TNF-a`*`IP-10`*`PDGF-AA`*`MDC`)^(1/4)),
         NF5 = ((`TNF-a`*`IP-10`*`MIP-1b`*`MDC`*`PDGF-AA`)^(1/5)))

### 7. Ploting correlation plot for different normalization factors calculated by different count of reference cetokines

cormat_control <- corr.test(nf[,10:15], y = NULL, use = "pairwise", method="spearman", adjust="BH", alpha = 0.05)

#tiff("corrplot_NF.tiff", units="in", width=7, height=7, res=800, family = "serif", pointsize = 14)
#jpeg("corrplot_NF.jpeg", units="in", width = 7, height = 7, res = 800, family = "serif", pointsize = 14)
corrplot::corrplot(as.matrix(cormat_control$r), type="full", method = 'square', tl.col = 'black', addCoef.col = "black", tl.cex= 0.7, cl.ratio = 0.2, tl.srt = 45, p.mat = cormat_control$p, insig = "blank", mar=c(0,0,1,0))
#dev.off()

# NF3 and NF4 have high correlation coefficient (~0.92), which means that after normalization to 3 reference or 4 reference cytokines, we will get similar results comparing the groups.
# And NF2 and NF3 have high correlation coefficient too (~0.81). 

# 8. 
y <- nf %>%
  gather(key = "Cytokine", value = "Concentration", - Sample, -Group, -Plate, -Standard,
         -NF_TNFa, -NF_IP10, -NF2, -NF3, -NF4, -NF5) %>%
 # mutate(Sample = as.character(Sample)) %>%
  dplyr::filter(!is.na(Concentration)) %>%
  arrange(Sample, Cytokine)

y$Concentration5 <- y$Concentration/y$NF5
y$Concentration4 <- y$Concentration/y$NF4
y$Concentration3 <- y$Concentration/y$NF3
y$Concentration2 <- y$Concentration/y$NF2
y$Concentration1 <- y$Concentration/y$NF_IP10
y$Concentration1.1 <- y$Concentration/y$NF_TNFa

y$Cytokine %>% unique()

stat.test5 <- compare_means(
  Concentration5 ~ Group, data = y, group.by = "Cytokine",
  method = "wilcox.test", paired = T, p.adjust.method = "BH", .id = "Sample") %>%
  select(Cytokine, `p.adj`) %>%
  dplyr::rename(`p.adj_NF5`=`p.adj`) 

stat.test4 <- compare_means(
  Concentration4 ~ Group, data = y, group.by = "Cytokine",
  method = "wilcox.test", paired = T, p.adjust.method = "BH", .id = "Sample") %>%
  select(Cytokine, `p.adj`) %>%
  dplyr::rename(`p.adj_NF4`=`p.adj`) 

stat.test3 <- compare_means(
  Concentration3 ~ Group, data = y, group.by = "Cytokine",
  method = "wilcox.test", paired = T, p.adjust.method = "BH", .id = "Sample") %>%
  select(Cytokine, `p.adj`) %>%
  dplyr::rename(`p.adj_NF3`=`p.adj`)

stat.test2 <- compare_means(
  Concentration2 ~ Group, data = y, group.by = "Cytokine",
  method = "wilcox.test", paired = T, p.adjust.method = "BH", .id = "Sample") %>%
  select(Cytokine, `p.adj`) %>%
  dplyr::rename(`p.adj_NF2`=`p.adj`)

stat.test1 <- compare_means(
  Concentration1 ~ Group, data = y, group.by = "Cytokine",
  method = "wilcox.test", paired = T, p.adjust.method = "BH", .id = "Sample") %>%
  select(Cytokine, `p.adj`) %>%
  dplyr::rename(`p.adj_NF_IP10`=`p.adj`)

stat.test1.1 <- compare_means(
  Concentration1.1 ~ Group, data = y, group.by = "Cytokine",
  method = "wilcox.test", paired = T, p.adjust.method = "BH", .id = "Sample") %>%
  select(Cytokine, `p.adj`) %>%
  dplyr::rename(`p.adj_NF_TNFa`=`p.adj`)

# В виде таблицы 
stat.test <- Reduce(function(...) merge(..., all = TRUE, by="Cytokine"),
                      list(stat.test1, stat.test1.1, stat.test2, stat.test3,
                           stat.test4, stat.test5))


stat.test

#write.xlsx(stat.test, "Different_NF.xlsx")

# *Note
# I compared cytokines levels between groups after normalization for different number of reference cytokines. And only concentration of 2 cytokines was equal after group comparison. (TNF-a and IP-10)
# So we decided to use only two control cytokines for normalization (TNF-a and IP-10).

```

### 9. Comparison the normalized concentration of cytokines btw groups by the paired Wilcoxon signed-ranks test

```{r}
# Merge imputated data with 'nf' dataset (included calculated NF)
colnames(data)[1:4] <- colnames(nf)[1:4]
prepared <- merge(data, nf[,c("Sample", "Group", "Plate", "Standard", "NF2")], by = c("Sample", "Group", "Plate", "Standard"))

# Normalize all cytokine concentration values
prepared[,5:15] <- sapply(prepared[,5:15], function(x) x/prepared$NF2)

# Compare the normalized logarithmic concentration of cytokines btw groups by paired wilcoxon
x <- prepared %>%
  gather(key = "Cytokine", value = "Concentration", - Sample, -Group, -Plate, -Standard) %>%
  mutate(Sample = as.character(Sample)) %>%
  dplyr::filter(!is.na(Concentration))

x <- x %>%
  dplyr::filter(Cytokine != "NF2") #exclude calculated NF from data 

#
stat1 <- x %>%
  group_by(Group, Cytokine) %>%
  dplyr::summarise(count = n(),
                   median = round(median(Concentration, na.rm = TRUE), digits = 2),
                   Q1 = round(quantile(Concentration, 0.25, na.rm = TRUE), digits = 2),
                   Q3 = round(quantile(Concentration, 0.75, na.rm = TRUE), digits = 2),
                   IQR = round(IQR(Concentration, na.rm = TRUE), digits = 2))

# 
Max <- x %>%
  group_by(Cytokine) %>%
  dplyr::summarise(`y.position` = max(Concentration, na.rm = TRUE)*1.08) # calculating y.position for plot

#
stat.test <- compare_means(
  Concentration ~ Group, data = x, group.by = "Cytokine",
  method = "wilcox.test", paired = T, .id = "Sample") 

stat.test <- stat.test[,-6] # remove p.adj calculated automatically 
stat.test

```

## 10. Visualize results of comparison

```{r, include=T}
stat.test <- stat.test %>%
  adjust_pvalue(p.col = "p", method = "BH") %>%
  add_significance("p.adj")

table_stat <- stat1 %>%
  select(Group, Cytokine, median, Q1, Q3) %>%
  mutate(DS = paste(median, " [", Q1, "; ", Q3, "] ")) %>%
  select(Group, Cytokine, DS) %>%
  cast(Cytokine ~ Group, value = "DS")

table_stat <- merge(table_stat, stat.test, by="Cytokine")

setwd("D:/Workflow/Atherosclerosis lab/Bloody plaques/Results/")
#write.xlsx(table_stat, "table_stat.xlsx")

stat.test <- merge(stat.test, Max, by="Cytokine")

colnames(x)

#tiff("Wilcox_cytokines.tiff", units="in", width=8, height=7, res=800, pointsize=6, family = "serif")
#jpeg("Wilcox_cytokines.jpeg", units="in", width = 8, height = 7, res = 800, family = "serif", pointsize = 6)
ggboxplot(x[x$Cytokine == "IP-10" | x$Cytokine == "TNF-a",], add = "jitter",
          x = "Group", y = "Concentration", fill = "Group", scales = "free", facet.by = "Cytokine", 
          width = 0.5, bxp.errorbar = T, size = 0.5, 
          palette = c("lightsalmon1", "powderblue")) +
  stat_pvalue_manual(stat.test[stat.test$Cytokine == "IP-10" | stat.test$Cytokine == "TNF-a",], label = "p.adj.signif", tip.length = 0.01) +
  theme(axis.text.x = element_blank(),
        text = element_text(family="serif", size = 14)) +
  ylab("Normalized cytokine level") +
  scale_y_continuous(labels = scales::comma)
#dev.off()

x$Cytokine %>% unique()

# tiff("Wilcox_reference_cytokines.tiff", units="in", width=12, height=12, res=800, pointsize=6, family = "serif")
#jpeg("Wilcox_reference_cytokines.jpeg", units="in", width = 12, height = 12, res = 800, family = "serif", pointsize = 6)
ggboxplot(x[x$Cytokine != "IP-10" & x$Cytokine != "TNF-a",], add = "jitter",
          x = "Group", y = "Concentration", fill = "Group", scales = "free", facet.by = "Cytokine", 
          width = 0.5, bxp.errorbar = T, size = 0.5, 
          palette = c("lightsalmon1", "powderblue")) +
  stat_pvalue_manual(stat.test[stat.test$Cytokine != "IP-10" & stat.test$Cytokine != "TNF-a",], label = "p.adj.signif", tip.length = 0.01) +
  theme(axis.text.x = element_blank(),
        text = element_text(family="serif", size = 14)) +
  ylab("Normalized cytokine level") +
  scale_y_continuous(labels = scales::comma)
#dev.off()

```

## 11. Normalize imputed values for other cytokines;
## 12. Comparing groups by signed rank test;

```{r}
colnames(data_st)[c(1,3,5)] <- c("Standard", "Sample", "Plate")
prepared <- merge(data_st, nf[,c("Sample", "Group", "Plate", "Standard", "NF2")], by = c("Sample", "Group", "Plate", "Standard"))

prepared <- prepared %>%
  mutate(Flag = ifelse(grepl("OOR", Concentration), "Imputed", "Measured"))

x <- prepared %>%
  mutate(Value = as.numeric(Value),
         `Normalized value` = Value/NF2)  

x <- x %>%
  mutate(Start_Interval = case_when(grepl("OOR >", Concentration) ~ `Normalized value`,
                                    grepl("OOR <", Concentration) ~ 0,
                                    !grepl("OOR", Concentration) ~ `Normalized value`),
         End_Interval = case_when(grepl("OOR >", Concentration) ~ 1000000,
                                  grepl("OOR <", Concentration) ~ `Normalized value`,
                                  !grepl("OOR", Concentration) ~ `Normalized value`))

colnames(x)
x$Group %>% unique()

Plasma <- x[x$Group == "Plasma", c("Sample", "Group",
               "Cytokine", "Concentration", "Imputation", "Normalized value",
               "Start_Interval", "End_Interval")]

Media <- x[x$Group == "Media", c("Sample", "Group",
               "Cytokine", "Concentration", "Imputation", "Normalized value",
               "Start_Interval", "End_Interval")]

colnames(Plasma)[c(2,4:8)] <- paste(colnames(Plasma)[c(2,4:8)], "Plasma", sep="_")
colnames(Media)[c(2,4:8)] <- paste(colnames(Media)[c(2,4:8)], "Media", sep="_")

Mergind_table <- merge(Plasma, Media, by=c("Sample", "Cytokine"))

colnames(Mergind_table)

Mergind_table <- Mergind_table %>%
  mutate(Result = case_when(Start_Interval_Media > End_Interval_Plasma ~ "Higher in Media",
                            Start_Interval_Plasma > End_Interval_Media ~ "Higher in Plasma",
                            Start_Interval_Media >= Start_Interval_Plasma & Start_Interval_Media <= End_Interval_Plasma ~ "Zero",
                            Start_Interval_Plasma >= Start_Interval_Media & Start_Interval_Plasma <= End_Interval_Media ~ "Zero",
                            Start_Interval_Media <= Start_Interval_Plasma & End_Interval_Plasma <= End_Interval_Media ~ "Zero",
                            Start_Interval_Media >= Start_Interval_Plasma & End_Interval_Plasma >= End_Interval_Media ~ "Zero"))

# Prepare tables for Signed rank test 

# Create table by cytokine with count of Result values

Signed_test <- Mergind_table %>%
  group_by(Cytokine, Result) %>%
  summarise(Count = n()) %>%
  cast(Cytokine~Result, value="Count") %>%
  replace(is.na(.), 0) %>%
  mutate(Sum = `Higher in Media` + `Higher in Plasma`)

Signed_test <- Signed_test 
Signed_test$p_value <- NA
Signed_test$p_value <- apply(Signed_test, 1, function(x) binom.test(c(as.numeric(x[2]), as.numeric(x[3])), as.numeric(x[5]), p=1/2, alternative = "two.sided")$p.value)

Signed_test <- Signed_test %>%
  adjust_pvalue(p.col = "p_value", method = "BH")

Signed_test

# write.xlsx(Signed_test, "Signed_rank_test.xlsx")
```

